\pdftitle

### 2018-05-15 maxNt_v3.tex
V3

%% Author: PGL  Porta Mana
%% Created: 2015-11-04T09:20:08+0100
%% Last-Updated: 2018-05-15T11:12:59+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Report-no: ***
\documentclass{article}

@@

%\newcommand*{\etal}{{et al.}}

@@

\enquote{Cooperativity} can quantified and characterized with \me\ method
in several ways. The simplest way proceeds along the following steps.
Consider the recorded activity of a sample of $n$ neurons.
\begin{enumerate}
\item The activity of each neuron, ideally a continuous signal, is
  transformed into into a sequence of \enquote{$0$}s (inactive) and
  \enquote{$1$}s (firing) by binning the time and binarizing the signal
  intensity \cite{caianiello1961,caianiello1986}.

@@

We hurry to say that the description just given is not \emph{the} way to
quantify and characterize cooperativity and functional grouping, but just
\emph{one} way. It can surely be criticized from many points of view. Yet,
it is quantitative and more precise than an undefined notion of
\enquote{cooperativity}. Two people who apply it to the same data will
obtain the same numbers. Other protocols can be based on the \me\ method,
for example protocols that take into account the activities or joint
activities of individual neurons, or even protocols that take into account
time dependence.

@@

To apply the \me\ method to the larger, unsampled population, it is
necessary to use probability relations relevant to sampling
\cites{ghoshetal1997}[parts~I,VI]{freedmanetal1978_r2007}[\chap~8]{gelmanetal1995_r2014}[\chap~3]{jaynes1994_r2003}.
The relations we present are well-known in survey sampling and in the
pedagogic problem of drawing from an urn without replacement, yet they are
somewhat hard to find explicitly written in the neuroscientific literature,
so they may be of interest on their own.

@@

\section{Setup: population, sample, probabilities}
\label{sec:setup}

We have a population of $N$ binary neurons. We assume that they can be
distinguished, by their spike shapes for example; but other details, like
their locations, are unknown. The neurons have a joint state
$(X_1,\dotsc,X_N) \defs \yX$ having fixed but unknown binary values
$(R_1,\dotsc,R_N)\defs \yR \in \set{0,1}^N$. A particular sample of $n$
neurons from this population has joint state $(x_1, \dotsc, x_n) \defs \yx$
having fixed binary values $(r_1, \dotsc, r_n)\defs\yr \in \set{0,1}^n$. We
will consider various averages of the population and the sample. For this
purpose we introduce a general averaging operator $\sav{\dotv}$ defined by
\begin{equation}
  \begin{gathered}
    \av{\yX} \defd \tfrac{1}{N} (X_1 + X_2 + \dotsb + X_N),
    % & \sav{\yx} &\defd \tfrac{1}{n} (x_1 + x_2 + \dotsb + x_n),
   \qquad% \\[2\jot]
    \av{\yX \yX} \defd \tbinom{N}{2}^{-1} (X_1 X_2 + X_1 X_3 + \dotsb +
    X_{N-1} X_N),
    % & \sav{\yx \yx} &\defd\tbinom{n}{2}^{-1} (x_1 x_2 + x_1 x_3 + \dotsb
    % + x_{n-1} x_n),
    \\[2\jot]
    \av{\yX\yX\yX} \defd \tbinom{N}{3}^{-1} (X_1 X_2 X_3 +% X_1 X_2 X_4
    \dotsb + X_{N-2} X_{N-1}X_N),
    % & \sav{\yx\yx\yx} &\defd \tbinom{n}{3}^{-1} (x_1 x_2 x_3
    % + %x_1 x_2 x_4
    % \dotsb + x_{n-2} x_{n-1} x_n),
  \end{gathered}
\end{equation}
and so on.
These formulae say that $\yXf$ is the fraction of active neurons, $\yXXf$
the fraction of simultaneously active pairs out of all $\binom{N}{2}$
pairs, $\av{\yX\yX\yX}$ the fraction of simultaneously active triplets, and
so on. Products of states like $X_i \dotsm X_j$ also have values in
$\set{0,1}$; from this we can combinatorially  prove that
\begin{equation}
  \label{eq:products_intermsof_average}
  \av{\underbrace{\yX\dotsm\yX}_{\text{$m$ factors}}}
  = \binom{N}{m}^{-1}\binom{N\yXf}{m}.
\end{equation}
Analogous formulae  hold for quantities like $\yx$, $\yR$, $\yr$.

@@

\section{Initial assumptions: the probability of representative samples}
\label{sec:prob_samples}

We need to make an initial probability assignment before any experimental
observations are
made. %, no matter what kinds of predictions we are interested in.
This initial assignment will be modified by our experimental observations.
Our probability assignment should reflect that the sample is somehow
\enquote{representative} of the population. We consider here two states of
knowledge that express this representativeness in different ways but lead
to identical \emph{initial} probability assignments.

In the first state of knowledge, denoted $\yHp$, we know that the neurons
in the population are biologically or functionally similar, for
example in morphology and kind of input or output they receive or give.
Knowledge of this similarity leads us to assign a probability distribution
for the population state $\yX$ that is symmetric under permutations of
neuron identities, or \emph{exchangeable} as it is usually called.

In the second state of knowledge or ignorance, denoted $\yHi$, we are
completely ignorant about the physical details of the individual neurons.
Our ignorance is therefore symmetric under permutations of neuron
identities. This also leads to an exchangeable probability distribution for
$\yX$.

Let us use $\yH$ to denote either of these two states of knowledge, in those
probabilities that are identical for $\yHp$ and $\yHi$.

The \emph{representation theorem for finite exchangeability} states that
the symmetric distribution of $\yH$ must obey
\begin{equation}
  \label{eq:joint_plaus_N_homog}
  \p(\yX = \yR \|  \yH) \equiv
  \p(\yX = \yR \| \yXf = \yRf, \yH)\,
  \p(\yXf=\yRf \| \yH) =
  \binom{N}{N\yRf}^{-1} \p(\yXf=\yRf \| \yH),
\end{equation}
the latter being the probability for the population average $\yX$. A sum is
only apparently missing in the central term: its summands $\yXf=A$ are all
zero except for $A=\yRf$. Proof of this theorem and generalizations to
non-binary and continuum cases are given by de~Finetti
\cite{definetti1959b}, Kendall \cite{kendall1967}, Ericson
\cite{ericson1976}, Diaconis \amp\ Freedman
\cite{diaconis1977,diaconisetal1980}, Heath \amp\ Sudderth
\cite{heathetal1976}. This theorem is intuitive: owing to symmetry, we must
assign equal probabilities to all states with $N\yRf$ active neurons.

>>>>

%% Author: PGL  Porta Mana
%% Created: 2015-11-04T09:20:08+0100
%% Last-Updated: 2018-05-15T12:04:44+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Report-no: ***
\documentclass{article}

@@

\newcommand*{\etal}{{et al.}}

@@

\enquote{Cooperativity} can quantified and characterized with \me\ method
in several ways. The simplest way proceeds along the following steps.
Consider the recorded activity of a sample of $n$ neurons.
\begin{enumerate}
\item The activity of each neuron, ideally a continuous signal, is
  transformed into a sequence of \enquote{$0$}s (inactive) and
  \enquote{$1$}s (firing) by binning the time and binarizing the signal
  intensity \cite{caianiello1961,caianiello1986}.

@@

We hurry to say that the description just given is just \emph{one} way to
quantify and characterize cooperativity and functional grouping, not
\emph{the} only way. It can surely be criticized from many points of view.
Yet, it is quantitative and more precise than an undefined notion of
\enquote{cooperativity}. Two people who apply it to the same data will
obtain the same numbers. Other protocols can be based on the \me\ method,
for example protocols that take into account the activities or joint
activities of individual neurons, or even protocols that take into account
time dependence.

@@

To apply the \me\ method to the larger, unsampled population, it is
necessary to use probability relations relevant to sampling
\cites{ghoshetal1997}[parts~I, VI]{freedmanetal1978_r2007}[\chap~8]{gelmanetal1995_r2014}[\chap~3]{jaynes1994_r2003}.
The relations we present are well-known in survey sampling and in the
pedagogic problem of drawing from an urn without replacement, yet they are
somewhat hard to find explicitly written in the neuroscientific literature,
so they may be of interest on their own.

@@

\section{Notation: population, sample, probabilities}
\label{sec:setup}

We have a population of $N$ binary neurons. We assume that they can be
distinguished, by their spike shapes for example; but other details, like
their locations, are unknown. The neurons have a joint state
$(X_1,\dotsc,X_N) \defs \yX$ having fixed but unknown binary values
$(R_1,\dotsc,R_N)\defs \yR \in \set{0,1}^N$; \enquote{$X$} denote the
variable, and \enquote{$R$} its value. A particular sample of $n$ neurons
from this population has joint state $(x_1, \dotsc, x_n) \defs \yx$ having
fixed binary values $(r_1, \dotsc, r_n)\defs\yr \in \set{0,1}^n$. We will
consider various averages of the population and the sample. For this
purpose we introduce a population-averaging operator $\sav{\dotv}$ defined
by
\begin{equation}
  \begin{gathered}
    \av{\yX} \defd \tfrac{1}{N} (X_1 + X_2 + \dotsb + X_N),
    % & \sav{\yx} &\defd \tfrac{1}{n} (x_1 + x_2 + \dotsb + x_n),
   \qquad% \\[2\jot]
    \av{\yX \yX} \defd \tbinom{N}{2}^{-1} (X_1 X_2 + X_1 X_3 + \dotsb +
    X_{N-1} X_N),
    % & \sav{\yx \yx} &\defd\tbinom{n}{2}^{-1} (x_1 x_2 + x_1 x_3 + \dotsb
    % + x_{n-1} x_n),
    \\[2\jot]
    \av{\yX\yX\yX} \defd \tbinom{N}{3}^{-1} (X_1 X_2 X_3 +% X_1 X_2 X_4
    \dotsb + X_{N-2} X_{N-1}X_N),
    % & \sav{\yx\yx\yx} &\defd \tbinom{n}{3}^{-1} (x_1 x_2 x_3
    % + %x_1 x_2 x_4
    % \dotsb + x_{n-2} x_{n-1} x_n),
  \end{gathered}
\end{equation}
and so on.
These formulae say that $\yXf$ is the fraction of active neurons, $\yXXf$
the fraction of simultaneously active pairs out of all $\binom{N}{2}$
pairs, $\av{\yX\yX\yX}$ the fraction of simultaneously active triplets, and
so on. Products of states like $X_i \dotsm X_j$ also have values in
$\set{0,1}$; from this we can combinatorially  prove that
\begin{equation}
  \label{eq:products_intermsof_average}
  \av{\underbrace{\yX\dotsm\yX}_{\text{$m$ factors}}}
  = \binom{N}{m}^{-1}\binom{N\yXf}{m}.
\end{equation}
Analogous formulae  hold for quantities like $\yx$, $\yR$, $\yr$.

@@

\section{Probability: relations between population and  sample}
\label{sec:prob_samples}

The theory of statistical sampling is covered in many excellent texts, for
example Ghosh \amp\ Meeden \citey{ghoshetal1997} or Freedman, Pisani, \amp\
Purves \citey[parts~I, VI]{freedmanetal1978_r2007}; summaries can be found
in Gelman \etal\ \citey[\chap~8]{gelmanetal1995_r2014} and Jaynes
\citey[\chap~3]{jaynes1994_r2003}.

We need to make an initial probability assignment for the state of the
population before any experimental observations are
made. %, no matter what kinds of predictions we are interested in.
This initial assignment will be modified by our experimental observations,
which can involve only a sample of the population. Our initial probability
assignment should reflect that samples are somehow representative of the
whole population. We consider here a state of knowledge that express
this representativeness.

In this state of knowledge, denoted $\yH$, we know that the neurons in the
population are biologically or functionally similar, for example in
morphology or the kind of input or output they receive or give. But we are
completely ignorant about the physical details of the individual neurons.
Our ignorance is therefore symmetric under permutations of neuron
identities. This ignorance leads us to assign a probability distribution
for the population state $\yX$ that is symmetric under permutations of
neuron identities; such a distribution is usually called \emph{finitely
  exchangeable} \cites{ericson1969}[\chap~1]{ghoshetal1997}. We stress that
this probability assignment is just an expression of the symmetry of our
\emph{ignorance} about the state of the population, not an expression of
some biologic or physical identity of the neurons.

The \emph{representation theorem for finite exchangeability} states that
the symmetric distribution in the state of knowledge $\yH$ must obey
\begin{equation}
  \label{eq:joint_plaus_N_homog}
  \p(\yX = \yR \|  \yH) \equiv
  \p(\yX = \yR \| \yXf = \yRf, \yH)\,
  \p(\yXf=\yRf \| \yH) =
  \binom{N}{N\yRf}^{-1} \p(\yXf=\yRf \| \yH),
\end{equation}
the latter being the probability for the population average $\yX$. A sum is
only apparently missing in the central term: its summands $\yXf=A$ are all
zero except for $A=\yRf$. Proof of this theorem and generalizations to
non-binary and continuum cases are given by de~Finetti
\cite{definetti1959b}, Kendall \cite{kendall1967}, Ericson
\cite{ericson1976}, Diaconis \amp\ Freedman
\cite{diaconis1977,diaconisetal1980}, Heath \amp\ Sudderth
\cite{heathetal1976}. This theorem is intuitive: owing to symmetry, we must
assign equal probabilities to all states with $N\yRf$ active neurons.
